---
title: "ML_final_coursework"
author: "Priyan Chandrapala"
date: "30/11/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Prediction Assignment Writeup
##Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

## Loading, cleaning and pre processing data
We start by loading the Libraries
```{r,echo=TRUE,warning=FALSE,message=FALSE}
rm(list=ls()) # remove old objects fro memory
setwd("//Users/priyanchandrapala/Documents/DataScienceCourse/09MachineLearning/week4")
library(knitr)
library(caret)
library(randomForest)
#library(AppliedPredictiveModeling)
library(doParallel)
```


```{r }
# download the files
#temp <- tempfile()
#download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",destfile='pml-training.csv', method='curl')
#download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",destfile='pml-test.csv', method='curl')
# load the data
pml_training <- read.csv("pml-training.csv",na.strings=c("NA","", " ","NULL","#DIV/0!"))
pml_testing <- read.csv("pml-test.csv",na.strings=c("NA","", " ","NULL","#DIV/0!"))
# check the dimensions of the data
dim(pml_training)
dim(pml_testing)
### Clean and minimise the data in the training set
# remove columns with mostly NA
pml_training_no_NA <- pml_training[ lapply( pml_training, function(x) sum(is.na(x)) / length(x) ) < 0.95 ]
dim(pml_training_no_NA)
# remove columns with near zero varience
NZV_columns <- nearZeroVar(pml_training_no_NA)
# remove near zero varience columns from the training set
pml_training_no_NA_no_NZV <- pml_training_no_NA[, -NZV_columns]
# find and remove corelated columns
# only numeric columns can be compaierd so extract the numeric columns 
num_cols<-sapply(pml_training_no_NA_no_NZV, is.numeric)
# extract the numeric columns
pml_training_numeric<-pml_training_no_NA_no_NZV[,num_cols]
# find the corelated columns
pml_training_numeric<-cor(pml_training_numeric)
# select columns which have corelation more then 80%
hc<-findCorrelation(pml_training_numeric,cutoff = .8)
hc<-sort(hc)
# cutoff columns which has corelation more then 80%
pml_training_reduced<-pml_training_no_NA_no_NZV[,-c(hc)]
dim(pml_training_reduced)
# remove identification variables, they are not useful for prediction
pml_training_reduced<-pml_training_reduced[,-c(1:5)]
# Make classe into a factor.
pml_training_reduced$classe<-as.factor(pml_training_reduced$classe)
# These are the column names we use for prediction
pml_training_column_names<-names(pml_training_reduced)
pml_training_column_names
# pre process the training variables by centering and scalling
pre_proc <- preProcess(pml_training_reduced)
# set seed to make results reproducible
set.seed(555)
pml_training_reduced<-predict(pre_proc,pml_training_reduced,with=FALSE)
dim(pml_training_reduced)
### now we are ready for prediction
```

## This section is for the predictions 
```{r }
# breakup the training dataset in to training and testing subsets 
inTrain  <- createDataPartition(pml_training_reduced$classe, p=0.7, list=FALSE)
trainSet <- pml_training_reduced[inTrain, ]
testSet  <- pml_training_reduced[-inTrain, ]
dim(trainSet)
dim(testSet)

#### The ML Models ########
# Create cluster for parallel execution, I found this approach to be faster with some of the models 
cl <- makeCluster(detectCores())
registerDoParallel(cl)

# 1. Support Vector Machine model 
# this model takes a very long time to train
control <- trainControl(method="repeatedcv", number=10, repeats=3,allowParallel= TRUE)
svm_Fit <- train(classe ~.,data=trainSet,method="svmRadial",trControl=control,verbose=FALSE)
# predict SVM training set
svm_predict_train <- predict(svm_Fit, trainSet)
confusionMatrix(svm_predict_train, trainSet$classe)
# predict SVM testing set
svm_predict_test <- predict(svm_Fit, testSet)
confusionMatrix(svm_predict_test, testSet$classe)

# 2. Random forest
# here I have kept the number of trees a low 200 because my laptop is low powered 
# and it takes a very long time to predict 
rm_fit <- train(classe ~.,data=trainSet,method="rf",trControl=control,verbose=FALSE)
# predict GLM training set
rf_predict_train <- predict(rm_fit, trainSet)
confusionMatrix(rf_predict_train, trainSet$classe)
# predict GLM testing set
rf_predict_test <- predict(rm_fit, testSet)
confusionMatrix(rf_predict_test, testSet$classe)

# 3. Gradient boosting 
gbm_fit <- train(classe ~.,data=trainSet,trControl=control, method="gbm",verbose=FALSE)
# predict GBM training set
gbm_predict_train <- predict(gbm_fit, trainSet)
confusionMatrix(gbm_predict_train, trainSet$classe)
# predict GBM testing set
gbm_predict_test <- predict(gbm_fit, testSet)
confusionMatrix(gbm_predict_test, testSet$classe)

results <- resamples(list(SVM=svm_Fit,RF=rm_fit,GBM=gbm_fit))
# summarize the distributions
summary(results)
# boxplots of results
bwplot(results)
# dot plots of results
dotplot(results)
# In addition to the above models I have also tried the Caret tree model but it gives me an error

```

From the graph above we can see that the Random Forest is the best model for this dataset 


```{r}
# once the best algorithm is selected retrain the model with the complete 
# training dataset. This is done because there will be better accuracy if the 
# training data set is as large as possible. 
final_model <- train(classe ~.,data=pml_training_reduced,method="rf",trControl=control,verbose=FALSE)


```

## Final model
```{r}
# Print and save
print(final_model)
print(final_model$finalModel)
# save the selected training model 
save(final_model, file="finalModel.RData")
```

OOB estimate of  error rate is less than 1%

## Predict on test data
```{r}
# apply the same data cleaning and pre processing steps on the test data
pml_training_column_names_minus_classe<-pml_training_column_names[!pml_training_column_names %in% c("classe")]
pml_testing_reduced<-pml_testing[,pml_training_column_names_minus_classe]
test_pre_proc <- predict(pre_proc, pml_testing_reduced,with=FALSE)
# do the prediction 
test_predictions <- predict(final_model, test_pre_proc)
stopCluster(cl)
# bind the prediction and problem_id together
results <- data.frame(problem_id=pml_testing$problem_id,prediction=test_predictions)
# results below
results

```

